\section{Methods}
\subsection{Curating the Stable Mutant Preference Dataset}\label{sec:dataset_curation}

We constructed the preference alignment dataset by combining experimental stability measurements from three complementary databases: MegaScale~\citep{tsuboyama2023mega}, Human Domainome 1~\citep{beltran2025site}, and FireProtDB~\citep{stourac2021fireprotdb}.
Each database contributes distinct coverage of the protein stability landscape.

\paragraph{Data sources.}
MegaScale provides large-scale protein folding stability measurements ($\Delta G$) across $\sim$100 protein domains obtained via a high-throughput proteolysis-based assay.
Human Domainome 1 provides site-saturation mutagenesis fitness measurements (growth rate-based) across 522 human protein domain libraries, totaling $\sim$563{,}000 single-point mutations.
FireProtDB is a manually curated database of thermodynamic stability data ($\Delta\Delta G$, $\Delta T_m$) linked to PDB structures, containing proteins with longer sequences (254--1{,}022 amino acids) than the other two sources.

\paragraph{Stability classification.}
For each protein domain, we classified single-point mutations as \textit{stable} or \textit{unstable} relative to the wild-type sequence.
A mutation was labeled stable if its measured fitness (or $\Delta G$) was greater than or equal to the wild-type value, and unstable otherwise.
For MegaScale, $\Delta G$ values reported as ``$>$5'' (outside the assay dynamic range) were treated as 5.0~kcal/mol.
Multi-site mutations (insertions, deletions, and double mutants) were excluded from the main training set; double mutants from MegaScale were retained separately for out-of-distribution evaluation.

\paragraph{Preference pair construction.}
For DPO training, each preference sample is a triplet $(x_{\text{wt}}, y_w, y_l)$ consisting of the wild-type sequence, a stable (preferred) mutant, and an unstable (dispreferred) mutant from the same protein domain.
Preference pairs were constructed by taking all pairwise combinations of stable and unstable mutations within each domain.
The stable and unstable mutations in a given pair need not occur at the same sequence position---the DPO objective operates on per-position log-likelihood ratios at each respective mutation site.

\paragraph{Train/validation split.}
We performed an intra-domain split to prevent information leakage: for each protein domain, 80\% of stable and unstable mutations were assigned to training and 10\% to validation, with preference pairs formed only within each split.
This ensures that no specific mutation appears in both training and validation sets, though both sets share the same set of 1{,}154 wild-type protein domains.

\paragraph{Sequence length filtering.}
All sequences were filtered to a maximum length of 256 amino acids to match the ESM-2 tokenizer input constraints used during training.
This filtering excluded the majority of FireProtDB entries (which contain longer sequences); these were instead reserved as an out-of-distribution benchmark for evaluating generalization to longer proteins (Section~\ref{sec:single_mutation_benchmark}).

\paragraph{Out-of-distribution holdouts.}
To evaluate in-distribution generalization at the protein family level, we held out specific Human Domainome domains from two PFAM families (PF00018 and PF07525) as out-of-distribution test sets.
These held-out domains share family membership with training domains but represent distinct proteins, enabling assessment of homolog-level generalization.

\paragraph{Final dataset statistics.}
The curated dataset comprises 1{,}154 unique protein domains spanning both MegaScale and Human Domainome sources.
The training set contains 184{,}938 unique stable mutants and 730{,}323 unique unstable mutants (915{,}261 total unique mutations), yielding 143{,}578{,}400 preference pairs.
The validation set contains 21{,}679 stable and 82{,}306 unstable unique mutants (103{,}985 total), yielding 1{,}824{,}750 preference pairs.
The out-of-distribution holdout from Human Domainome PF00018 and PF07525 families contains 421{,}641 preference pairs.
Only single-point mutations were used for training; the complete dataset statistics and curation pipeline are summarized in Figure~\ref{fig:intro_stableESM}c.

\subsection{Implementing StableESM: LoRA-based Direct Preference Optimization of ESM2}\label{sec:stableesm_implementation}

StableESM applies Low-Rank Adaptation (LoRA) to pretrained ESM-2 models, inserting trainable low-rank matrices into the feed-forward layers while keeping all original parameters frozen.
We used LoRA rank $r = 32$, scaling factor $\alpha = 64$, and dropout $0.1$, targeting the \texttt{fc1} and \texttt{fc2} modules (the two linear layers in each transformer block's feed-forward network).
No bias terms were added to the LoRA layers.
This configuration was held constant across all model sizes (8M, 35M, 150M, 650M, and 3B parameters), enabling a controlled comparison of how preference alignment scales with model capacity.

\paragraph{Reference model.}
DPO requires a reference model to regularize the aligned model against degenerate solutions.
We initialize the reference model as a deep copy of the full model (including LoRA parameters at initialization) and freeze all of its parameters.
During training, forward passes through the reference model are performed under \texttt{torch.no\_grad()} to avoid unnecessary gradient computation and reduce memory overhead.

\paragraph{Masked log-likelihood ratios.}
Our DPO formulation operates on per-position log-likelihood ratios (LLRs) computed via masked prediction, following the zero-shot mutation scoring approach from ESM~\citep{meier2021language}.
For a mutation at position $i$, we mask position $i$ in the input sequence and compute:
\begin{equation}
\pi(y \mid x) = \frac{p(x_i = \text{mut} \mid x_{\setminus i})}{p(x_i = \text{wt} \mid x_{\setminus i})}
\end{equation}
where $x_{\setminus i}$ denotes the sequence with position $i$ replaced by the \texttt{<mask>} token.
Probabilities are obtained by applying a softmax over the full vocabulary at the masked position and indexing the specific wild-type and mutant amino acid tokens.
The mutation position is offset by $+1$ during token indexing to account for the ESM-2 \texttt{<cls>} token prepended to all sequences.

\paragraph{DPO loss.}
Each training sample consists of a triplet $(x_{\text{wt}}, y_w, y_l)$: wild-type sequence, a stable (preferred) mutation, and an unstable (dispreferred) mutation from the same protein domain.
The stable and unstable mutations may occur at different sequence positions.
For each mutation, we compute the LLR under both the policy model $\pi_\theta$ and the frozen reference model $\pi_{\text{ref}}$.
The DPO objective is:
\begin{equation}
\mathcal{L}_{\text{DPO}} = -\mathbb{E}_{(x, y_w, y_l) \sim \mathcal{D}} \left[ \log \sigma \left( \beta \log \frac{\pi_\theta(y_w \mid x)}{\pi_{\text{ref}}(y_w \mid x)} - \beta \log \frac{\pi_\theta(y_l \mid x)}{\pi_{\text{ref}}(y_l \mid x)} \right) \right]
\end{equation}
where $\beta = 0.8$ controls the strength of the KL divergence constraint against the reference model.
This objective trains the model to increase the relative likelihood of stable mutations over unstable mutations, while the reference model term prevents the aligned model from diverging too far from the pretrained representations.

\paragraph{Tokenization and masking.}
Sequences are tokenized using the ESM-2 batch converter.
Stop codons (\texttt{*}) are replaced with the \texttt{<eos>} token.
All sequences are padded to a maximum length of 256 tokens.
For each preference triplet, we construct seven tokenized variants: the unmasked wild-type, stable, and unstable sequences, plus four masked versions (wild-type masked at the stable mutation position, wild-type masked at the unstable mutation position, stable mutant masked at its mutation position, and unstable mutant masked at its mutation position).
These masked sequences enable extraction of the amino acid probabilities needed for the LLR computation under both policy and reference models.

\subsection{Training StableESM across model sizes using Direct Preference Optimization}\label{sec:dpo_training}

We trained StableESM models at five scales using the ESM-2 family: 8M (\texttt{esm2\_t6\_8M\_UR50D}), 35M (\texttt{esm2\_t12\_35M\_UR50D}), 150M (\texttt{esm2\_t30\_150M\_UR50D}), 650M (\texttt{esm2\_t33\_650M\_UR50D}), and 3B (\texttt{esm2\_t36\_3B\_UR50D}).
All models were trained on the same preference dataset (Section~\ref{sec:dataset_curation}) using the DPO objective described in Section~\ref{sec:stableesm_implementation}.

\paragraph{Optimizer and learning rate schedule.}
We used the Adam optimizer with a two-phase learning rate schedule implemented as a \texttt{SequentialLR} scheduler.
The first phase applies a linear warmup from $0.1 \times \text{lr}$ to the peak learning rate over 1{,}000 steps.
The second phase follows a cosine annealing schedule with period $T_{\max} = 100{,}000$ steps and minimum learning rate set to $0.01 \times \text{lr}$.
The scheduler updates every step.

\paragraph{Model-specific hyperparameters.}
Learning rates were scaled inversely with model size to maintain training stability.
Batch sizes and distributed training strategies were adjusted to fit GPU memory constraints.
Table~\ref{tab:training_hyperparams} summarizes the training configuration for each model size.

\begin{table}[h]
\centering
\caption{Training hyperparameters for StableESM models across scales.}
\label{tab:training_hyperparams}
\small
\begin{tabular}{lccccc}
\toprule
 & \textbf{8M} & \textbf{35M} & \textbf{150M} & \textbf{650M} & \textbf{3B} \\
\midrule
Learning rate & $1 \times 10^{-6}$ & $5 \times 10^{-7}$ & $1 \times 10^{-7}$ & $5 \times 10^{-8}$ & $1 \times 10^{-8}$ \\
Batch size & 32 & 32 & 8 & 8 & 4 \\
Max steps & 8M & 4M & 4M & 4M & 4M \\
Precision & bf16 & bf16 & bf16 & bf16 & bf16-mixed \\
Strategy & DDP & DDP & DDP & DeepSpeed S2 & DeepSpeed S2 \\
GPUs & 1 & 1 & 4 & 4 & 8 (2 nodes) \\
$\beta$ (DPO) & 0.8 & 0.8 & 0.8 & 0.8 & 0.8 \\
LoRA rank $r$ & 32 & 32 & 32 & 32 & 32 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Training infrastructure and precision.}
All models were trained using PyTorch Lightning.
Smaller models (8M, 35M) used Distributed Data Parallel (DDP) on a single GPU, while larger models used multi-GPU setups: 150M used DDP across 4 GPUs, 650M used DeepSpeed Stage 2 across 4 GPUs, and 3B used DeepSpeed Stage 2 across 8 GPUs on 2 nodes.
Training used bfloat16 (bf16) mixed precision for memory efficiency on A100 GPUs.
Validation was performed every 10{,}000 steps, computing Spearman correlation per protein domain averaged across all domains.
Data loading used 4 workers per GPU.
Gradient accumulation was not used (accumulation factor of 1).

\paragraph{Monitoring convergence.}
During training, we tracked the DPO loss, the probability that the model prefers the stable mutation over the unstable mutation ($P(\text{stable} > \text{unstable})$), and per-class accuracy (fraction of stable mutations assigned higher likelihood than wild-type, and fraction of unstable mutations assigned lower likelihood than wild-type).
All model sizes showed smooth convergence with $P(\text{stable} > \text{unstable})$ increasing from the random baseline of 0.5 to approximately 0.70, with aligned training and validation trajectories indicating no overfitting (Figure~\ref{fig:train_eval_stableesm}a).

\subsection{Zero-shot mutation effect prediction using log-likelihood ratio}\label{sec:llr_prediction}

StableESM predicts mutational effects at inference time using a reference-normalized log-likelihood ratio (LLR) that requires no task-specific prediction head.
For a single mutation at position $i$ that substitutes wild-type amino acid $a_{\text{wt}}$ with mutant amino acid $a_{\text{mut}}$, the score is:
\begin{equation}\label{eq:llr_inference}
\text{LLR}(i) = \log \frac{p_\theta(a_{\text{mut}} \mid x_{\setminus i})}{p_\theta(a_{\text{wt}} \mid x_{\setminus i})} - \log \frac{p_{\text{ref}}(a_{\text{mut}} \mid x_{\setminus i})}{p_{\text{ref}}(a_{\text{wt}} \mid x_{\setminus i})}
\end{equation}
where $p_\theta$ is the aligned StableESM model, $p_{\text{ref}}$ is the frozen ESM-2 reference model, and $x_{\setminus i}$ denotes the sequence with position $i$ replaced by the \texttt{<mask>} token.
The first term captures how much the aligned model favors the mutation; the second term subtracts the base model's evolutionary prior, isolating the effect of preference alignment.

\paragraph{Scoring procedure.}
For each mutation, we perform four masked forward passes: the wild-type and mutant sequences are each passed through both the policy and reference models with the mutation position masked.
At the masked position, we apply a softmax over the full vocabulary and extract the probabilities for the wild-type and mutant amino acid tokens.
These four probabilities are combined via Equation~\ref{eq:llr_inference} to produce the final score.
A positive LLR indicates a predicted stabilizing mutation; a negative LLR indicates a destabilizing one.

\paragraph{Multi-site mutations.}
For mutations at multiple positions (\eg double mutants), we use a marginal scoring approach that sums the individual per-position LLR contributions:
\begin{equation}
\text{LLR}_{\text{multi}} = \sum_{j} \text{LLR}(j)
\end{equation}
where each position $j$ is independently masked and scored.
This assumes additivity of mutational effects---each position is scored with the other positions unmasked, providing the native sequence context at all non-mutated sites.

\paragraph{Comparison to vanilla ESM-2 scoring.}
For the base ESM-2 model (without preference alignment), the standard masked marginal scoring from \citet{meier2021language} computes $\text{LLR}_{\text{ESM2}} = \log p(a_{\text{mut}} \mid x_{\setminus i}) - \log p(a_{\text{wt}} \mid x_{\setminus i})$ using only the pretrained model.
StableESM's reference-normalized formulation (Equation~\ref{eq:llr_inference}) can be interpreted as measuring the \textit{change} in mutational preference induced by stability alignment, on top of the evolutionary baseline captured by the reference model.

\subsection{Benchmarking StableESM on single mutations}\label{sec:single_mutation_benchmark}

We evaluated StableESM on single-point mutation effect prediction across four progressively stringent generalization settings: (1) held-out mutations from training domains, (2) held-out homolog domains from training families, (3) held-out protein families, and (4) completely out-of-distribution long-sequence proteins.

\paragraph{In-distribution validation (held-out mutations).}
For each of the 1{,}154 training domains, 10\% of stable and unstable mutations were held out during training (Section~\ref{sec:dataset_curation}).
At validation, we computed LLR scores (Equation~\ref{eq:llr_inference}) for all held-out mutations within each domain and calculated the Spearman rank correlation ($\rho$) between predicted LLR scores and experimental fitness values ($\Delta G$ for MegaScale, growth rate for Human Domainome).
Correlations were computed per domain and then averaged across all domains to produce a mean Spearman $\rho$.
Validation was performed every 10{,}000 training steps.

\paragraph{Out-of-distribution homolog domains.}
To test generalization to unseen proteins within represented families, we evaluated on held-out Human Domainome domains from the PF00018 and PF07525 PFAM families.
These domains share family membership with training proteins but represent distinct sequences.
For each held-out domain, we performed a full deep mutational scan---scoring all 19 possible amino acid substitutions at every position---and computed the Spearman $\rho$ between predicted LLR scores and experimental fitness values.
We also computed the area under the ROC curve (AUC) for binary classification of stabilizing versus destabilizing mutations (using LLR $\geq 0$ as the positive prediction threshold).

\paragraph{Out-of-distribution protein families (FireProtDB).}
FireProtDB proteins were excluded from training due to their sequence lengths (254--1{,}022 amino acids) exceeding the 256-token training cutoff.
This provides a stringent out-of-distribution benchmark testing generalization to both unseen protein families and substantially longer sequences.
For each FireProtDB protein, we computed LLR scores for all available mutations and calculated Spearman $\rho$ against experimental $\Delta\Delta G$ and $\Delta T_m$ values separately.
Performance was reported as the per-protein correlation averaged across all FireProtDB proteins, with the fraction of proteins showing improved correlation over ESM-2 (win rate) as an additional metric.

\paragraph{Evaluation across model scales.}
All benchmarks were evaluated for ESM-2, StableESM, and ESM$_{\text{Therm}}$ (supervised fine-tuning baseline) at 8M, 35M, 150M, 650M, and 3B parameters.
ESM$_{\text{Therm}}$ models were trained using the original ESM$_{\text{Therm}}$ training code and datasets with a supervised stability predictor head, enabling a controlled comparison between DPO-based alignment and conventional supervised fine-tuning.
For ESM$_{\text{Therm}}$, mutation effect scores were computed using the standard masked marginal LLR from \citet{meier2021language} (without reference normalization), since these models do not use a reference model framework.

\subsection{Benchmarking StableESM on double mutations}\label{sec:double_mutation_benchmark}

To test whether preference alignment on single mutations transfers to higher-order mutational effects, we evaluated StableESM on double mutation stability prediction using MegaScale data.

\paragraph{Dataset.}
Double mutations were identified in the MegaScale dataset by the presence of colon-delimited mutation specifications (\eg \texttt{D1A:K2R}) and separated from the single-mutation training data during curation.
The double mutation benchmark comprises 190 unique protein domains, each with experimental $\Delta G$ measurements for multiple double mutants.
These proteins were not used for DPO training, which included only single-point mutations.

\paragraph{Scoring.}
Double mutations were scored using simultaneous masking: both mutation positions were masked at once, and log-softmax probabilities were extracted at each masked position independently.
The total LLR for a double mutant at positions $i$ and $j$ is the sum of per-position contributions:
\begin{equation}
\text{LLR}_{\text{double}} = \text{LLR}(i) + \text{LLR}(j)
\end{equation}
where each $\text{LLR}(k)$ is the reference-normalized log-likelihood ratio (Equation~\ref{eq:llr_inference}) computed at position $k$ with both mutation positions masked.
This assumes approximate additivity of mutational effects while providing each position with the same masked context.

\paragraph{Evaluation.}
For each of the 190 proteins, we computed the Spearman rank correlation between predicted LLR scores and experimental $\Delta G$ values.
Performance was assessed by comparing the distributions of per-protein Spearman correlations between StableESM and ESM-2, using the Kolmogorov--Smirnov test for statistical significance and Cohen's $d$ for effect size.
All five model sizes (8M, 35M, 150M, 650M, 3B) were evaluated.

\subsection{Benchmarking StableESM on disorder region and solubility prediction}\label{sec:disorder_prediction}

To evaluate whether stability-aligned representations transfer to protein properties unrelated to the training objective, we benchmarked StableESM on two tasks: per-residue disorder region prediction and sequence-level solubility classification.
Neither task was targeted during DPO training, so any improvements reflect more generally informative representations rather than task-specific memorization.
Both tasks use a common linear probing framework that isolates representation quality by freezing all backbone parameters and training only a single linear layer.

\paragraph{Linear probing framework.}
For each backbone, we froze all parameters and trained a single linear layer ($\text{Linear}(d_{\text{hidden}}, 1)$) as the prediction head.
The hidden dimension $d_{\text{hidden}}$ matches the backbone's representation size: 320 (8M), 480 (35M), 640 (150M), 1{,}280 (650M), and 2{,}560 (3B).
For the one-hot baseline, input dimensionality is 33 (vocabulary size) for disorder and $1{,}024 \times 33 = 33{,}792$ (flattened) for solubility.
All probes were trained using AdamW with learning rate $1 \times 10^{-4}$, weight decay $1 \times 10^{-4}$, binary cross-entropy loss, maximum sequence length of 1{,}024 tokens, and FP16 mixed precision for up to 1{,}000 epochs.
Best models were selected by validation loss.

\paragraph{Backbones compared.}
We evaluated five backbones across model sizes: (1) StableESM (DPO-aligned policy model), (2) ESM-2 (unmodified pretrained baseline), (3) ESM$_{\text{Therm}}$ (supervised fine-tuning with stability predictor head), (4) one-hot encoding (minimal sequence baseline), and (5) randomly initialized ESM-2 weights (untrained control).
This set of controls distinguishes improvements from learned representations versus trivial sequence features or model capacity effects.

\subsubsection{Disorder region prediction}

\paragraph{Datasets.}
We trained on the DisProt training set and evaluated on the CAID (Critical Assessment of Intrinsic Disorder) benchmark test set comprising 584 proteins.
Per-residue disorder labels were provided as binary annotations aligned to sequence positions.

\paragraph{Architecture and training.}
The linear head was applied per residue, producing a disorder probability at each position.
Binary cross-entropy loss was computed with attention masking to exclude padding tokens from the loss and metric computation.
Training used a batch size of 32.

\paragraph{Metrics.}
Performance was quantified using per-protein F1 score, accuracy, precision, and recall, computed over active (non-padding) residues only and aggregated across all 584 CAID benchmark proteins.

\subsubsection{Solubility prediction}

\paragraph{Dataset.}
We used a binary solubility classification dataset with predefined train, validation, and test splits.
Each protein was labeled as soluble or insoluble.
Sequences longer than 1{,}022 amino acids were excluded to fit within the 1{,}024-token limit (accounting for \texttt{<cls>} and \texttt{<eos>} tokens).

\paragraph{Architecture and training.}
The linear head was applied to a mean-pooled representation over all non-padding tokens, producing a single sequence-level prediction.
Training used a batch size of 128.

\paragraph{Metrics.}
Performance was quantified using AUROC, F1 score, accuracy, precision, and recall on the held-out test set.

\subsection{Testing catastrophic forgetting on structure prediction}\label{sec:structure_forgetting}

To assess whether StableESM retains the structural knowledge encoded during pretraining, we evaluated its representations using the structure module of ESMFold~\citep{lin2023evolutionary}.
Embeddings and attention maps from StableESM were passed directly into the ESMFold structure module (trained for 270k update steps), following the evaluation protocol from the original ESMFold study.
The structure module was kept frozen throughout all experiments and no recycling steps were performed during prediction.

We evaluated models at 8M, 35M, 150M, 650M, and 3B parameters.
For each model size, embeddings and attention tensors were extracted from the same layers and attention heads as in the original ESMFold implementation.
Structural predictions were generated for protein targets from the CAMEO, CASP14, CASP15, and CASP16 benchmark datasets~\citep{yuan2025casp16}.
Performance was quantified using the average per-sequence pLDDT score.

To compare against conventional fine-tuning, we additionally evaluated representations from ESM$_{\text{Therm}}$ models~\citep{chu2024protein}.
ESM$_{\text{Therm}}$ models were trained with supervised fine-tuning using a regression head on stability data and the original ESM$_{\text{Therm}}$ training code and datasets.
We evaluated ESM$_{\text{Therm}}$ at matching parameter sizes, extracting embeddings and attention tensors from the same layers as for ESM-2 and StableESM.
This enables a controlled comparison of how DPO-based alignment versus supervised fine-tuning affects retention of structural signal.

\subsection{Testing catastrophic forgetting on clinical variant prediction}\label{sec:clinical_forgetting}

To evaluate whether stability alignment compromises pathogenicity prediction, we benchmarked StableESM on the ProteinGym clinical variant benchmark~\citep{notin2023proteingym}, which contains curated ClinVar pathogenic and benign substitutions across human proteins.

\paragraph{Scoring.}
Variants were scored using the same zero-shot LLR approach as for mutation effect prediction (Equation~\ref{eq:llr_inference}).
For each protein, pathogenic variants are expected to have negative LLR scores (predicted destabilizing or deleterious) and benign variants positive or near-zero scores.

\paragraph{Evaluation.}
We evaluated ESM-2, StableESM, and ESM$_{\text{Therm}}$ at 8M, 35M, 150M, 650M, and 3B parameters.
Performance was quantified per protein using AUC (area under the ROC curve for discriminating pathogenic from benign variants) and F1 score.
Aggregate performance was computed as the average across all proteins in the benchmark.
We additionally compared against TransceptEVE-L as an external state-of-the-art baseline.

\paragraph{Per-protein analysis.}
Beyond aggregate metrics, we examined the per-protein F1 score distribution by comparing StableESM against ESM-2 on a scatter plot.
This analysis reveals whether stability alignment uniformly preserves clinical prediction or differentially affects specific proteins---for instance, proteins where pathogenicity is driven by destabilization may benefit from stability alignment, while proteins with gain-of-function disease mechanisms may not.

\subsection{Testing catastrophic forgetting on ProteinGym benchmarks}\label{sec:proteingym_forgetting}

To assess whether preference alignment preserves general fitness prediction capabilities, we benchmarked StableESM on the ProteinGym substitution benchmark~\citep{notin2023proteingym}, which spans five functional categories: fitness, activity, binding, expression, and stability.

\paragraph{Scoring.}
Fitness scores were computed using the zero-shot LLR approach (Equation~\ref{eq:llr_inference}) for StableESM and the standard masked marginal LLR from \citet{meier2021language} for ESM-2.
For ESM$_{\text{Therm}}$, we evaluated both the supervised regression head (on stability only) and the LLR-based zero-shot approach to isolate representation quality from task-specific head performance.

\paragraph{Evaluation.}
We evaluated ESM-2, StableESM, and ESM$_{\text{Therm}}$ at 8M, 35M, 150M, 650M, and 3B parameters.
For each model, predicted scores were compared against experimentally measured fitness values across all ProteinGym deep mutational scanning datasets within each category.
Performance was quantified using Spearman’s rank correlation ($\rho$), computed per dataset and averaged within each functional category.
Results were compared against state-of-the-art models reported on the ProteinGym leaderboard.

\subsection{Designing multicopper oxidase alleles and variants with StableESM}\label{sec:oxidase_design}
To validate StableESM predictions, we experimentally tested designed mutations in laccase (CotA) from \textit{Bacillus subtilis} (UniProtKB accession P07788), a protein family not represented in the preference alignment dataset. We selected mutations based on StableESM predictions using the machine learning-directed evolution framework described in Algorithm~\ref{alg:ml_directed_evolution}, including overlap variants where both StableESM and ESM2 predict stabilizing effects and StableESM-specific predictions. Enzyme activity was assessed through ABTS oxidation assays at multiple temperatures (25$^\circ$C, 65$^\circ$C, and 85$^\circ$C). All experiments were performed in quadruplicate.

For large-scale in silico design validation, we applied machine learning-directed evolution on multicopper oxidase (PDB: 1GSK), a 513 amino acid protein not represented in the preference alignment dataset. We evaluated three design strategies: (1) StableESM greedy sampling, selecting positions with highest predicted stabilizing effects; (2) StableESM unique sampling, targeting positions where StableESM predicts stabilizing mutations but ESM2 does not (LLR$_{\text{StableESM}} \geq 0$ and LLR$_{\text{ESM2}} < 0$); and (3) ESM2 greedy sampling as baseline. Thermal stability was evaluated using TempBERTura~\citep{rodella2024temberture}, and structural quality was assessed using ColabFold with mean pLDDT scores.

\subsection{Predicting mutational effects with physics-based methods: Rosetta}\label{sec:rosetta_methods}

\subsection{Experimental protocol for protein purification}\label{sec:protein_purification}

\subsubsection{Cloning and preparing DNA constructs}
Laccase (CotA) from \textit{Bacillus subtilis} (UniProtKB accession P07788) was codon-optimized for \textit{Escherichia coli}, synthesized as a gBlock (Twist Bioscience, South San Francisco, CA, USA), and cloned into pY71 vector using Gibson Assembly. Site-directed mutations (A296G, G270C, I368V, and T261F) were introduced using the Q5\textsuperscript{\textregistered} Site-Directed Mutagenesis Kit (NEB, E0554S) with pY71-Laccase as the template. All constructs were verified by DNA sequencing.

\subsubsection{Cell-free expression and protein quantification}
Cell-free expression (CFE) reactions were performed using the PUREfrex2.0 system at 30$^\circ$C for 20 hours, and the concentration of the expressed protein was determined by Bradford assay (ThermoFisher). Protein standards (0, 0.25, 0.5, 1, 1.5, 2 mg/mL) were prepared and used as the standard curve. The absorbance at 595 nm was measured after mixing 4 $\mu$L of protein standards with 200 $\mu$L of Bradford reagent and allowing it to react at room temperature for 20 min. The protein concentration in each CFE reaction was determined from the standard curve, using no DNA as a blank control. Each sample was measured in four replicates.

\subsubsection{Copper activation and heat treatment}
Since laccase is activated by copper, we added copper sulfate (CuSO$_4$) into the CFE reaction after 20 hours of incubation. Specifically, 5 mM of CuSO$_4$ was added to the CFE system to reach a final concentration of 400 $\mu$M and incubated at room temperature (25$^\circ$C) for 2 hours in a pH 5.5 buffer environment (50 mM PIPES, 20 mM NaCl). After copper activation, each laccase mutant was subjected to heat treatment at three temperatures (85$^\circ$C, 65$^\circ$C, 25$^\circ$C) for 10 minutes to assess thermostability.

\subsection{Experimental protocol for activity assays}\label{sec:activity_assays}
The multicopper oxidase activity was measured using a colorimetric assay of 2,2'-azino-bis(3-ethylbenzothiazoline-6-sulfonic acid) (ABTS) oxidation, which turns blue upon oxidation. Colorimetric assays were performed in a pH 5.5 buffer environment (50 mM PIPES, 20 mM NaCl) containing 4 mM ABTS. The absorbance at 420 nm was measured once per minute for 30 minutes. Each sample was measured in four replicates.